Understanding the Landscape of Childlessness in the U.S.

### 2024-03-18 outputs/paper/paper.qmd
finalize paper

---
title: "Understanding the Landscape of Childlessness in the U.S."
Subtitle: "An analysis of General Social Survey data (1972-2022)"
author: 
  - Mehrnoush Mohammadi 
  - Quang Mai
thanks: "Code and data are available at: https://github.com/mehrnoush68/childlessness-in-america.git"
date: today
date-format: long
abstract: "We investigate U.S. childlessness trends between 1972 and 2022 using data from the General Social Survey. The research focuses on the relationship between childlessness and factors such as age, gender, education, social class, health conditions, marital status, and number of children. We find rising childlessness across demographics, potentially linked to economic factors, education, and social norms. We explore the social and economic implications of this trend and emphasize the need for further research on childlessness."
format: pdf
number-sections: true
toc: true
bibliography: references.bib
---

@@

#### Load data ####
table_nochild_age <- read_csv(file = "../data/clean_data_age.csv",show_col_types = FALSE)
table_nochild_class <- read_csv(file = "../data/clean_data_class.csv",show_col_types = FALSE)
table_nochild_degree <- read_csv(file = "../data/clean_data_degree.csv",show_col_types = FALSE)
table_nochild_sex <- read_csv(file = "../data/clean_data_sex.csv",show_col_types = FALSE)
table_nochild_marital <- read_csv(file = "../data/clean_data_marital.csv",show_col_types = FALSE)
table_nochild_health <- read_csv(file = "../data/clean_data_health.csv",show_col_types = FALSE)

@@

This offers a window into understanding not only the prevalence of childlessness within different segments of the population but also the potential drivers behind this trend. Existing research suggests a complex interplay of factors, including economic considerations, rising educational attainment, changing social norms related to career and family life, and individual preferences [@voluntaryChild].

@@

Structurally, in [Introduction], our paper starts with a brief dissection of the nature of childlessness in America and the contributing factors. Subsequently, in [Data] and [Result], we talk about the nature of the data obtained and analyze the results garnered from the data with suitable tables and charts. Next, [Discussion] provides further insights and future areas of study. Finally, [Conclusion] summarizes our main findings. To complete the paper, [Appendix] clarifies how each variable within each dataset is measured.

@@

@@

The 2022 iteration of the GSS, for the first time in its history, incorporated a combination of in-person, web self-administered, and phone interview methods. This expansion in data collection methods offers a potentially richer and more diverse sample. The GSS team is transparent about these limitations and welcomes feedback from the research community to continuously improve their methodology.

@@

@@

@fig-nochild-age shows a consistent rise in childlessness across all age groups in the United States from 1972 to 2022. It is important to acknowledge, however, that the data series has gaps for several years within this timeframe (1979, 1981, 1992, 1995, 1997, 1999, 2001, 2003, 2005, 2007, 2009, 2011, 2013, 2015, 2017, and 2019). These missing data points can potentially limit the generalization of the observed trends and hinder a comprehensive understanding of childlessness over time.
Despite these limitations, the available data reveals a steady increase in the percentage of individuals without children across all age groups. Notably, the most significant rise is observed among younger adults (18-34 and 35-49). As an example, the percentage of childless individuals in the 18-34 age group increased by approximately 42.70% from 1972 to 2022.

```{r}
#| echo: false
#| include: true
#| warning: false
#| label: fig-nochild-age
#| fig-cap: "Percentage of Childless Adults by Age Group in the US (1972-2020)"

@@

ggplot(data_long_age, aes(x = as.integer(year), y = percentage, color = Age)) +
  geom_point() +
  geom_smooth(method = "loess", se = FALSE, linetype = "solid", size = 1.2) +  # Add running average trend line
  labs(x = "Year", y = "Percentage of Population", title = "Childless Percentage of Population over Time by Age Group", color = "Age Group") +
  theme_minimal() +
  theme(text = element_text(size = 10))

@@

## Exploring Socioeconomic Class and Educational Attainment
To gain a deeper understanding of the factors influencing these trends, we look into additional data sets from NORC [@gss] examining the relationship between childlessness and socioeconomic class @tbl-nochild-class and educational attainment @tbl-nochild-degree.

@@

@tbl-nochild-class reveals a potential link between socioeconomic class and childlessness. The data suggests that individuals in the lower and working classes exhibit higher percentages of childlessness compared to those in the middle and upper classes. This pattern might be attributed to various factors, such as financial constraints, job insecurity, or limited access to childcare resources experienced by individuals in lower socioeconomic classes.

@@

@fig-nochild-class depicts the gradual increase in the percentage of childless individuals across four social classes (lower class, working class, middle class, and upper class) in the United States from 1972 to 2022. The graph reveals an upward trend in childlessness for all classes over the five decades.

```{r}
#| echo: false
#| include: true
#| warning: false
#| label: fig-nochild-class
#| fig-cap: "Percentage Childless over Time by Class"

# Plot the data
ggplot(data_long_class, aes(x = Year, y = Percentage, color = `Class Identification`)) +
  geom_point() +
  geom_smooth(method = "loess", se = FALSE, linetype = "solid", size = 1.2) +  # Add running average trend line
  labs(x = "Year", y = "Percentage of Population", title = "Childless Percentage of Population over Time by Class", color = "Breakdown Category") +
  theme_minimal() +
  theme(text = element_text(size = 10))

@@

While all classes show an increase in @fig-nochild-class, the data suggests the most significant rise occurred among the lower class, with their percentage of childless individuals nearly doubling from approximately 17.16% in 1972 to around 34.46% in 2022, marking a substantial increase of approximately 101.04% over the specified period. The working class also experienced a significant increase, from around 21.83% in 1972 to approximately 31.44% in 2022, indicating a rise of roughly 43.98%. Additionally, the middle class saw a slight increase, from approximately 28.24% in 1972 to about 30.09% in 2022, representing a rise of approximately 6.55%. Similarly, the upper class saw an increase, with their percentage of childless individuals climbing from approximately 20.24% in 1972 to about 27.87% in 2022, indicating an increase of roughly 37.71%.

@tbl-nochild-degree explores the association between educational attainment and childlessness. Interestingly, the data indicates a nuanced relationship. While individuals with higher educational degrees (college+) display a slightly higher percentage of childlessness compared to those with lower degrees, the difference is not as substantial as observed with socioeconomic class. This suggests that educational attainment might play a complex role in childbearing decisions, potentially influenced by factors like career aspirations, financial considerations, and personal preferences.

\newpage
```{r}
#| echo: false
#| include: true
#| warning: false
#| label: tbl-nochild-degree

@@

@fig-nochild-degree presents a comparative analysis of childlessness across individuals with different levels of educational attainment (less than high school, high school and college degree or higher) in the United States from 1972 to 2022. The graph utilizes line trends to depict the percentage of childless individuals within each degree category over the five decades.

The data reveals a distinct pattern across educational groups. Individuals with a high school diploma consistently exhibit the highest percentage of childlessness throughout the entire period, ranging from around 28.15% in 1972 to 34.56% in 2022. Conversely, the percentage of childless individuals with a college and higher degree has decreased from 32.85% in 1972 to 29.84% in 2022. The trend for individuals with a with less than a high school education increased from 16.97% in 1972 to 22.90% in 2022.
```{r}
#| echo: false
#| include: true
#| warning: false
#| label: fig-nochild-degree
#| fig-cap: "Percentage Childless over Time by Level of Education"

@fig-nochild-gender explores gender differences in childlessness within the United States over a 50-year period (1972-2022). The graph utilizes line trends to depict the percentage of childless individuals categorized by gender (Male and Female).

```{r}
#| echo: false
#| include: true
#| warning: false
#| label: fig-nochild-gender
#| fig-cap: "Percentage Childless over Time by Gender"

```
The data reveals a gradual increase in childlessness for both genders over the past five decades. However, females consistently exhibit a lower percentage of childlessness compared to males throughout the entire period. While the gap appears to be slightly narrowing over time, females still have a lower percentage by around 8 percentage points in 2022 (approximately 28.32% for females and 36.09% for males).

@@

@tbl-nochild-health uncovers the relationship between health conditions and childlessness. Intriguingly, those with 'excellent' and 'good' health are substantially more likely to be childless than those with 'fair' and 'poor' health. Several factors are attributable to this fact, but foremost, career, lifestyle differences and financial considerations might come into play. However, more data is needed to clarify the context of this trend, especially coupling it with variables like educational attainment or socioeconomic class. In addition, sampling bias might factor into this trend, where it's assumable that people with excellent health are often the younger population and those with less desirable health are older.

@@

@fig-nochild-health explores health differences in childlessness within the United States over a 50-year period (1972-2022). The graph utilizes trend lines to depict the percentage of childless individuals categorized by health conditions (excellent, good, fair, and poor).

Overall, the data shows an increase in childlessness across all health levels and conditions approaching 2022. As mentioned, people with excellent health have the highest percentage of childlessness, having a 10.49% increase throughout the 50-year period (28.26% in 1972 to 38.75% in 2022). People with good and fair health exhibit a similar rate of childlessness in 2022, at about 30%. In particular, people with fair health marked; however, people with poor health, once having the lowest rate of childlessness (at 12.27% in 1972), now increases to 20.40% in 2022, which translates to a substantial 8.13% increase.  
```{r}
#| echo: false
#| include: true
#| warning: false
#| label: fig-nochild-health
#| fig-cap: "Percentage Childless over Time by Health Conditions"

@tbl-nochild-marital explores people's marital status and childlessness. Trends of childlessness vary with this dataset. Unsurprisingly, people who are never married have the highest percentage of childlessness. However, those who are divorced, widowed and separated have fluctuating percentages of childlessness, which is understandable due to the complex emotional nature of these marital statuses.

@@

@fig-nochild-marital explores the marital status of people who are childless over a span of 50 years in the United States (1972-2022). The graph contains trend lines of varying marital statuses-Divorced, Married, Never married, Separated and Widowed-and how each has changed over the years in terms of childbearing status.

The data demonstrates that a significant portion of people who have never married are childless. However, more and more people are reluctant to get married but still consider having children. This is observable in the childlessness percentage decrease of people who never marry, decreasing from 96.73% in 1972 to 72.53% in 2022. For people who are divorced, separated and widowed, this percentage fluctuates, displaying the complex dynamics between these relationship statuses and childbearing. For married people though, this childlessness percentage also decreases, from 10.51% in 1972 to 13.97% in 2022.

```{r}
#| echo: false
#| include: true
#| warning: false
#| label: fig-nochild-marital
#| fig-cap: "Percentage Childless over Time by Marital Status"

# Plot the data
ggplot(data_long_marital, aes(x = Year, y = Percentage, color = `Marital status`)) +
  geom_point() +
  geom_smooth(method = "loess", se = FALSE, linetype = "solid", size = 1.2) +  # Add running average trend line
  labs(x = "Year", y = "Percentage of Population", title = "Childless Percentage of Population over Time by Marital Status", color = "Marital status") +
  theme_minimal() +
  theme(text = element_text(size = 10))

The figures presented in this paper were generated using data from the [@gss]. These figures visually represent the key findings related to the trends and patterns in childlessness across different ages, social classes, educational attainment levels, and genders in the United States over the past five decades (1972-2022). It is important to remember that these figures solely present descriptive data and do not establish causal relationships. The Discussion section will further analyze the observed trends and explore the broader implications of these findings.

@@

### Incomplete Data for Certain Years
While the General Social Survey provides valuable data on childlessness trends, there are gaps in data availability for certain years. This creates a limitation in our ability to present a completely continuous picture of trends across the entire 50-year time frame.

@@

```{r}
#| echo: false
#| include: true
#| warning: false
#| label: fig-nochild-avg-marital
#| fig-cap: "Average Percentage Childless over Time by Marital Status"

@@

# Plot the data
ggplot(avg_data_marital, aes(x = Year, y = avg_Percentage)) +
  geom_line() +
  geom_point() +
  labs(x = "Year", y = "Average Percentage of Population", title = "Average Childless Percentage of Population over Time by Marital Status") +
  theme_minimal() +
  theme(text = element_text(size = 9))

@@

```{r}
#| echo: false
#| include: true
#| warning: false
#| label: fig-nochild-avg-class
#| fig-cap: "Average Percentage Childless over Time by Class"

@@

# Plot the data
ggplot(avg_data_class, aes(x = Year, y = avg_Percentage)) +
  geom_line() +
  geom_point() +
  labs(x = "Year", y = "Average Percentage of Population", title = "Average Percentage Childless over Time by Class") +
  theme_minimal() +
  theme(text = element_text(size = 9))

```{r}
#| echo: false
#| include: true
#| warning: false
#| label: fig-nochild-avg-gender
#| fig-cap: "Average Percentage Childless over Time by Gender"

# Group the data by Year and calculate the average of Percentage for each year
avg_data_sex <- data_long_sex |>
  group_by(Year) |>
  summarise(Sum_Percentage = sum(Percentage, na.rm = TRUE)) |>
  mutate(avg_Percentage = Sum_Percentage / 2)

# Print the average data
# print(avg_data_sex)

# Plot the data
ggplot(avg_data_sex, aes(x = Year, y = avg_Percentage)) +
  geom_line() +
  geom_point() +
  labs(x = "Year", y = "Average Percentage of Population", title = "Average Percentage Childless over Time by Gender") +
  theme_minimal() +
  theme(text = element_text(size = 9))

```{r}
#| echo: false
#| include: true
#| warning: false
#| label: fig-nochild-avg-degree
#| fig-cap: "Average Percentage Childless over Time by Degree"

@@

# Plot the data
ggplot(avg_data_degree, aes(x = Year, y = avg_Percentage)) +
  geom_line() +
  geom_point() +
  labs(x = "Year", y = "Average Percentage of Population", title = "Average Percentage Childless over Time by Degree") +
  theme_minimal() +
  theme(text = element_text(size = 9))

@@

```{r}
#| echo: false
#| include: true
#| warning: false
#| label: fig-nochild-avg-degree-class
#| fig-cap: "Average Percentage Childless over Time by Degree and Class"

# Merge the two datasets based on the Year column
merged_data <- merge(avg_data_degree, avg_data_class, by = "Year", all = TRUE, suffixes = c("_degree", "_class"))

# Calculate the sum of avg_Percentage for each year
merged_data <- merged_data |>
  mutate(sum_Percentage = (coalesce(avg_Percentage_degree, 0) + coalesce(avg_Percentage_class, 0))/2)

# Print the merged and summed data
# print(merged_data)

# Plot the merged and summed data
ggplot(merged_data, aes(x = Year, y = sum_Percentage)) +
  geom_line() +
  geom_point() +
  labs(x = "Year", y = "Sum of Average Percentage of Population", title = "Sum of Average Percentage Childless over Time by Degree and Class") +
  theme_minimal() +
  theme(text = element_text(size = 9))
```

@@

## Rising Childlessness: Patterns and Trends
The findings presented in this paper reveal a consistent and concerning trend of rising childlessness across various demographics in the United States. As @fig-nochild-age illustrates, the percentage of individuals without children has steadily increased across all age groups over the past 50 years, with the most significant rise observed among younger individuals (18-34 and 35-49).

@@

This paper cannot definitively establish causal relationships, but the data suggests several potential factors contributing to rising childlessness, all supported by existing research. These include economic concerns like rising childcare costs, student loan debt, and economic uncertainty. Additionally, individuals with higher educational attainment, particularly women, might delay or forgo childbearing to pursue career opportunities. Furthermore, shifting social norms related to childbearing, family structures, and career aspirations, especially for women, may be contributing factors. Finally, the decision to remain childless can also be influenced by individual values, lifestyle choices, and personal circumstances.

@@

The rising trend of childlessness carries potential consequences across various societal spheres. A shrinking pool of younger workers could pose challenges to economic growth and strain social security systems. Additionally, the rise of childless families could have implications for social support networks and intergenerational relationships. While childlessness can be a personal choice for some individuals and may even lead to a fulfilling life, it could also be associated with feelings of isolation and a lack of social support for others. These potential consequences highlight the need for further research and discussion on how to navigate the social, economic, and individual complexities associated with rising childlessness.

>>>>

---
title: "Understanding the Landscape of Childlessness in the U.S."
subtitle: "An analysis of General Social Survey data (1972-2022)"
author: 
  - Mehrnoush Mohammadi 
  - Quang Mai
thanks: "Code and data are available at: https://github.com/mehrnoush68/childlessness-in-america.git"
date: today
date-format: long
abstract: "We investigate U.S. childlessness trends between 1972 and 2022 using data from the General Social Survey. The research focuses on the relationship between childlessness and factors such as age, gender, education, social class, health conditions, marital status, and number of children. We find rising childlessness across demographics, potentially linked to economic factors, education, and social norms. We explore the social and economic implications of this trend and emphasize the need for further research on childlessness."
format: pdf
number-sections: true
toc: true
bibliography: references.bib
---

@@

#### Load data ####
table_nochild_age <- read_csv(file = "../data/clean_data_age.csv",show_col_types = FALSE)
table_nochild_class <- read_csv(file = "../data/clean_data_class.csv",show_col_types = FALSE)
table_nochild_degree <- read_csv(file = "../data/clean_data_degree.csv",show_col_types = FALSE)
table_nochild_sex <- read_csv(file = "../data/clean_data_sex.csv",show_col_types = FALSE)
table_nochild_marital <- read_csv(file = "../data/clean_data_marital.csv",show_col_types = FALSE)
table_nochild_health <- read_csv(file = "../data/clean_data_health.csv",show_col_types = FALSE)
table_nochild_overall <- read_csv(file =
"../data/nochild_overall.csv",show_col_types = FALSE)

table_nochild_overall <- table_nochild_overall |>
  drop_na() |>
  mutate_at(vars(-1), as.numeric)
```

@@

These helps us understand not only the prevalence of childlessness within different segments of the population but also the potential drivers behind this trend. Existing research suggests a complex interplay of factors, including economic considerations, rising educational attainment, changing social norms related to career and family life, and individual preferences [@voluntaryChild].

@@

Our estimand is the relationship between age, gender, socio-economic class, educational attainment, health status, marital status and the rate of childlessness in the United States from 1972 to 2022. Through our analysis, we found that childless rate is consistently on the rise from 1972 to 2022, with correlation coefficient of 0.73 between childless percentage and time (or a 6.99% increase from 1972 to 2022). Factors that directly contribute to this rise are higher educational attainment, higher socio-economic class,lower gender disparity, better health status and dissolution of marriages through divorces, separation and being widowed.

To further understand why these causes explain rising childlessness, our paper starts with a brief dissection of the nature of childlessness in America and the contributing factors in [Introduction]. Subsequently, in [Data] and [Result], we talk about the nature of the data obtained and analyze the results garnered from the data with suitable tables and charts. Next, [Discussion] provides further insights and future areas of study. Finally, [Conclusion] summarizes our main findings. To complete the paper, [Appendix] clarifies how each variable within each dataset is measured and the survey that we've generated based off the dataset.

@@

## Measurement

Each variable used—age, class, educational attainment, health, marital status, and gender—is  quantified through direct questioning to accurately capture respondents' backgrounds. For instance, the General Social Survey (GSS) obtains information on marital status through questions such as, "Are you currently married, widowed, divorced, separated, or have you never been married?", with responses categorized into predefined groups for further analysis [@gssCodebook]. On ther other hand, instead of relying on income sources, socio-economic class is subjectively assessed based on the respondents' answers to this question, "If you were asked to use one of four names for your social class, which would you say you belong in: the lower class, the working class, the middle class, or the upper class?". Some are continuous variable, like age. With respondents' various age groups documented, we can understand the generational disparities in childbearing patterns. Systematic and data-driven, these measurement methods ensures that all data faithfully reflects the American population's childbearing tendencies.

@@

The 2022 iteration of the GSS, for the first time in its history, incorporated a combination of in-person, web self-administered, and phone interview methods in response to COVID-19. This expansion in data collection methods offers a potentially richer and more diverse sample. As a result, after implementing these changes, the GSS randomly assigned respondents to various survey modes to examine response patterns and data accuracy. Unsurprisingly, compared to last year's 17%, 94.2% of participants opted for a new survey mode, which is a significant increase of 50.5% in the overall response rate. This emphasizes the effectiveness of employing multimodal survey strategies to boost survey participation.

However, this multimodal survey approach also presents several challenges in collecting respondents' opinions on topics such as childbirth and childbearing tendencies, given the private nature of these subjects. With the easy formatting and non-required nature of these surveys, biases like acquiescence bias (only positive responses) might occur because respondents aren't confronted in a face-to-face settings, where many will just opt a neutral response [@webSurvey]. Therefore, we carefully select various variables to to ensure the authenticity of the public's background on the topic of childlessness, avoiding having the data being too influenced by the changes in surveying methods.

## Exploring Age and Generational Differences

@@

Thus, there is a consistent rise in childlessness across all age groups in the United States from 1972 to 2022. Notably, the most significant rise is observed among younger adults (18-34 and 35-49). It is important to acknowledge, however, that the data series has gaps for several years within this timeframe (1979, 1981, 1992, 1995, 1997, 1999, 2001, 2003, 2005, 2007, 2009, 2011, 2013, 2015, 2017, and 2019). These missing data points can potentially limit the generalization of the observed trends and hinder a comprehensive understanding of childlessness over time.

@@

```{r}
#| echo: false
#| include: false
#| warning: false
#| label: fig-nochild-age
#| fig-cap: "Childless Percentage in the US by Age Group (1972-2020)"

@@

ggplot(data_long_age, aes(x = as.integer(year), y = percentage, color = Age)) +
   geom_line() +
  geom_smooth(method = "loess", se = FALSE, linetype = "dashed", size = 1.2) +  # Add running average trend line
  labs(x = "Year", y = "Percentage of Population", title = "Childless Percentage of Population over Time by Age Group", color = "Age Group") +
  theme_minimal() +
  theme(text = element_text(size = 10))

@@

The data reveals a gradual increase in childlessness for both genders over the past five decades. However, females consistently exhibit a lower percentage of childlessness compared to males throughout the entire period. While the gap appears to be slightly narrowing over time, females still have a lower percentage by around 8 percentage points in 2022 (approximately 28.32% for females and 36.09% for males).

```{r}
#| echo: false
#| include: false
#| warning: false
#| label: fig-nochild-gender
#| fig-cap: "Childless Percentage in the US over Time by Gender (1972-2020)"

## Exploring Socioeconomic Class and Educational Attainment
To gain a deeper understanding of the factors influencing these trends, we look into additional data sets from NORC [@gss] examining the relationship between childlessness and socioeconomic class (@tbl-nochild-class) and educational attainment (@tbl-nochild-degree).

```{r}
#| include: true
#| label: fig-nochild-class-edu
#| fig-cap: Childless Percentage in the US by Class and Educational Attainment (1972-2020)
#| fig-subcap: [Childless percentage by class, Childless percentage by education]
#| layout-ncol: 2
#| warning: false
#| message: false
#| echo: false
#| fig-pos: "t"

# Plot the data
ggplot(data_long_class, aes(x = Year, y = Percentage, color = `Class Identification`)) +
   geom_line() +
  geom_smooth(method = "loess", se = FALSE, linetype = "dashed", size = 1.2) +  # Add running average trend line
  labs(x = "Year", y = "Percentage of Population", title = "Childless Percentage of Population over Time by Class", color = "Breakdown Category") +
  theme_minimal() +
  theme(text = element_text(size = 10))

@@

### Overview of Socioeconomic Class's Impact on Childlessness

@@

@tbl-nochild-class reveals a potential link between socioeconomic class and childlessness. To visualize the data, @fig-nochild-class-edu depicts the gradual increase in the percentage of childless individuals across four social classes (lower class, working class, middle class, and upper class) in the United States from 1972 to 2022. The graph reveals an upward trend in childlessness for all classes over the five decades.

@@

### Overview of Educational Attainment's Impact on Childlessness

Further, @tbl-nochild-degree explores the association between educational attainment and childlessness. @fig-nochild-class-edu illustrates this data, presenting a comparative analysis of childlessness across individuals with different levels of educational attainment (less than high school, high school and college degree or higher) in the United States from 1972 to 2022. The graph utilizes line trends to depict the percentage of childless individuals within each degree category over the five decades.

```{r}
#| echo: false
#| include: true
#| warning: false
#| label: tbl-nochild-degree

@@

@@

```{r}
#| include: true
#| label: fig-nochild-health-marital
#| fig-cap: Childless Percentage in the US over Time by Marital and Health Status (1972-2020)
#| fig-subcap: [Childless percentage by health, Childless percentage by marital status]
#| layout-ncol: 2
#| warning: false
#| message: false
#| echo: false
#| fig-pos: "t"

# Plot the data
ggplot(data_long_marital, aes(x = Year, y = Percentage, color = `Marital status`)) +
  geom_line() +
  geom_smooth(method = "loess", se = FALSE, linetype = "dashed", size = 1.2) +  # Add running average trend line
  labs(x = "Year", y = "Percentage of Population", title = "Childless Percentage of Population over Time by Marital Status", color = "Marital status") +
  theme_minimal() +
  theme(text = element_text(size = 10))

\newpage
### Overview of Health Status's Impact on Childlessness

@tbl-nochild-health uncovers the relationship between health conditions and childlessness. Accordingly, the first graph of @fig-nochild-health-marital explores the health differences in people who are childless within the United States over a 50-year period (1972-2022). The graph utilizes trend lines to depict the percentage of childless individuals categorized by health conditions (excellent, good, fair, and poor).

@@

### Overview of Marital Status's Impact on Childlessness

@tbl-nochild-marital explores people's marital status and childlessness. @fig-nochild-health-marital then illustrates this data, exploring the marital status of people who are childless over a span of 50 years in the United States (1972-2022). The graph contains trend lines of varying marital statuses-Divorced, Married, Never married, Separated and Widowed-and how each has changed over the years in terms of childbearing status.

@@

@@

### Incomplete Data for Certain Years
While the General Social Survey provides valuable data on childlessness trends, there are gaps in data availability for certain years. In summary, out of 68596 survey entrees, there are 250 missing cases. This creates a limitation in our ability to present a completely continuous picture of trends across the entire 50-year time frame.

@@

## Health Conditions and Marital Status Impact on Childlessness

Overall, the data shows an increase in childlessness across all health levels and conditions approaching 2020. As mentioned, @fig-nochild-health-marital shows that people with excellent health have the highest percentage of childlessness, having a 10.49% increase throughout the 50-year period (28.26% in 1972 to 38.75% in 2022). People with good and fair health exhibit a similar rate of childlessness in 2022, at about 30%. However, people with poor health, once having the lowest rate of childlessness (at 12.27% in 1972), now increases to 20.40% in 2022, which translates to a substantial 8.13% increase. Overall, on average, across all health levels from 1972-2022, @fig-avg-one shows that percentage of childlessness increases at about 6.836%.

Unintuitively, those with 'excellent' and 'good' health are substantially more likely to be childless than those with 'fair' and 'poor' health. Several factors are attributable to this fact, but foremost, career, lifestyle differences and financial considerations might come into play. However, more data is needed to clarify the context of this trend, especially coupling it with variables like educational attainment or socioeconomic class. In addition, sampling bias might factor into this trend, where it's assumable that people with excellent health are often the younger population and those with less desirable health are older.

As for marital status, @fig-nochild-health-marital demonstrates that a significant portion of people who have never married are childless. However, more and more people are reluctant to get married but still have children. This is observable in the childlessness percentage decrease of people who never marry, decreasing from 96.73% in 1972 to 72.53% in 2022. For people who are divorced, separated and widowed, this percentage is on the rise fluctuates, displaying the complex dynamics between these relationship statuses and childbearing. For married people though, this childlessness percentage also decreases, from 10.51% in 1972 to 13.97% in 2022. Overall, on average, across all marital status from 1972-2022, @fig-avg-one shows that percentage of childlessness decreases at about 7.39%.

Trends of childlessness vary with this dataset. Unsurprisingly, people who are never married have the highest percentage of childlessness. However, those who are divorced, widowed and separated have fluctuating but similarly increasing percentages of childlessness, which is understandable due to the complex emotional nature of these marital statuses. The overall trend of people who are in a relationship or are married having more children (7.39% decrease of childlessness) are attributed to the higher quality of life, and other factors.

```{r}
#| include: true
#| label: fig-avg-one
#| fig-cap: Average Childless Percentage per Year in the US (1972-2022)
#| fig-subcap: [Childless percentage by health, Childless percentage by marital status]
#| layout-ncol: 2
#| warning: false
#| message: false
#| echo: false
#| fig-pos: "t"

# Group the data by Year and calculate the average of Percentage for each year
avg_data_health <- data_long_health |>
  group_by(Year) |>
  summarise(Sum_Percentage = sum(Percentage, na.rm = TRUE)) |>
  mutate(avg_Percentage = Sum_Percentage / 5)

# Print the average data
# print(avg_data_health)

# Plot the data
ggplot(avg_data_health, aes(x = Year, y = avg_Percentage)) +
  geom_line() +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "blue") +
  labs(x = "Year", y = "Average Percentage per Year") +
  theme_minimal() +
  theme(text = element_text(size = 9))

@@

# Plot the data
ggplot(avg_data_marital, aes(x = Year, y = avg_Percentage)) +
  geom_line() +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "blue") +
  labs(x = "Year", y = "Average Percentage per Year") +
  theme_minimal() +
  theme(text = element_text(size = 9))

@@

## Between Socio-economic Class, Educational Attainment and Childlessness

While all classes show an increase of childlessness in @fig-nochild-class-edu, the data suggests the most significant rise occurred among the lower class, with their percentage of childless individuals nearly doubling from approximately 17.16% in 1972 to around 34.46% in 2022, marking a substantial increase of approximately 101.04% over the specified period. The working class also experienced a significant increase, from around 21.83% in 1972 to approximately 31.44% in 2022, indicating a rise of roughly 43.98%. Additionally, the middle class saw a slight increase, from approximately 28.24% in 1972 to about 30.09% in 2022, representing a rise of approximately 6.55%. Similarly, the upper class saw an increase, with their percentage of childless individuals climbing from approximately 20.24% in 1972 to about 27.87% in 2022, indicating an increase of roughly 37.71%.

The data suggests that individuals in the lower and working classes exhibit higher percentages of childlessness compared to those in the middle and upper classes. This pattern might be attributed to various factors, such as financial constraints, job insecurity, or limited access to childcare resources experienced by individuals in lower socioeconomic classes.

Similarly, data reveals a distinct pattern across educational groups. Individuals with a high school diploma consistently exhibit the highest percentage of childlessness throughout the entire period, ranging from around 28.15% in 1972 to 34.56% in 2022. Conversely, the percentage of childless individuals with a college and higher degree has decreased from 32.85% in 1972 to 29.84% in 2022. The trend for individuals with a with less than a high school education increased from 16.97% in 1972 to 22.90% in 2022.

Thus, the data indicates a nuanced relationship. While individuals with higher educational degrees (college+) display a slightly higher percentage of childlessness compared to those with lower degrees, the difference is not as substantial as observed with socioeconomic class. This suggests that educational attainment might play a complex role in childbearing decisions, potentially influenced by factors like career aspirations, financial considerations, and personal preferences.

```{r}
#| include: true
#| label: fig-avg-two
#| fig-cap: Average Childless Percentage per Year in the US (1972-2022)
#| fig-subcap: [Childless percentage by class, Childless percentage by degree]
#| layout-ncol: 2
#| warning: false
#| message: false
#| echo: false
#| fig-pos: "t"

@@

# Plot the data
ggplot(avg_data_class, aes(x = Year, y = avg_Percentage)) +
  geom_line() +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "blue") +
  labs(x = "Year", y = "Average Percentage per Year") +
  theme_minimal() +
  theme(text = element_text(size = 9))

@@

# Plot the data
ggplot(avg_data_degree, aes(x = Year, y = avg_Percentage)) +
  geom_line() +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "blue") +
  labs(x = "Year", y = "Average Percentage per Year") +
  theme_minimal() +
  theme(text = element_text(size = 9))

@@

## Overall Rate of Childlessness Over Time

```{r}
#| include: true
#| label: fig-overall
#| fig-cap: Overall Childless Percentage per Year in the US (1972-2022)
#| warning: false
#| message: false
#| echo: false
#| fig-pos: "t"

data_long_overall <- pivot_longer(table_nochild_overall, -`Year`, names_to = "year", values_to = "Percentage")

data_long_overall <- data_long_overall |>
select(-`Year`) |>
rename(`Year` = `year`) |>
arrange(`Year`) |>
mutate_all(as.numeric)

# Print the average data
# print(data_long_overall)

# Plot the data
data_long_overall$Year <- as.numeric(data_long_overall$Year)
data_long_overall$Percentage <- as.numeric(data_long_overall$Percentage)

# Plot the data with points, line, and trend line
ggplot(data_long_overall, aes(x = Year, y = Percentage)) +
  geom_point() +
  geom_line() +  # Add a line connecting the points
  geom_smooth(method = "lm", se = FALSE, color = "blue") +  # Add a linear trend line
  labs(x = "Year", y = "Overall Percentage per Year") +
  theme_minimal() +
  theme(text = element_text(size = 9)) +
  scale_x_continuous(labels = function(x) as.character(x))  # Convert numeric labels to character if needed

```{r}
#| include: false
# Calculate the correlation coefficient
correlation <- cor(data_long_overall$Year, data_long_overall$Percentage)

# Print the correlation coefficient
print(paste("Correlation coefficient between Year and Percentage:", correlation))
```

\newpage
@fig-overall graphs a trend-line of overall childlessness rate throughout the years. With a correlation coefficient between "Year" and "Percentage" of 0.73, it's highly likely that as time progresses, childlessness is on the rise. This is evident in the 6.99% increase in childlessness from 1972 to 2022.

To account for the general increase of childlessness from 1972 to 2022, there are several crucial events. First, in @fig-overall, the first surge of childlessness occurs from 1986 onwards till 1900 (27.71% to 30.26%, a 2.55% increase). This was when the AIDs pandemic occurred, where there was much fear and health concerns of HIV transmission in having sexual intercourse. The second surge of childlessness is observable from 2000 onwards till 2006 (27.76% to 28.94%, a 1.18% increase). This was the time period marked by fear of terrorism when 9/11 happened, where heightened feelings of vulnerability may have influenced people's decisions of expanding their family. The third surge of childlessness from 2008 till 2014 is drastic due to the occurrence of the Great Recession (27.25% to 30.75%, a 3% increase). Financial instability then is a contributing factor in determining childlessness. The last surge occured from 2018 to 2021 when the COVID-19 was in place (29.69% to 31.67%, a 1.98% increase).

@@

## Rising Childlessness: Patterns and Trends
The findings presented in this paper reveal a consistent and concerning trend of rising childlessness across various demographics in the United States. As @fig-overall illustrates, the percentage of individuals without children has steadily increased across all age groups over the past 50 years, with the most significant rise observed among younger individuals (18-34 and 35-49).

@@

This paper cannot definitively establish causal relationships, but the data suggests several potential factors contributing to rising childlessness, all supported by existing research. These include economic concerns like two global pandemics, rising childcare costs, student loan debt, and economic uncertainty [@voluntaryChild]. Additionally, individuals with higher educational attainment, particularly women, might delay or forgo childbearing to pursue career opportunities [@effectiveVol]. Furthermore, shifting social norms related to childbearing, family structures, and career aspirations, especially for women, may be contributing factors. Finally, the decision to remain childless can also be influenced by individual values, lifestyle choices, and personal circumstances.

@@

The rising trend of childlessness carries potential consequences across various societal spheres. A shrinking pool of younger workers could pose challenges to economic growth and strain social security systems. Additionally, the rise of childless families could have implications for social support networks and intergenerational relationships [@voluntaryChild]. While childlessness can be a personal choice for some individuals and may even lead to a fulfilling life, it could also be associated with feelings of isolation and a lack of social support for others [@effectiveVol]. These potential consequences highlight the need for further research and discussion on how to navigate the social, economic, and individual complexities associated with rising childlessness.
